---
title: "meaRtools: Reading in MEA data"
author: "Stephen J Eglen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{meaRtools: Reading in MEA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r rsetup-knitr,eval=TRUE,include=FALSE}
require(knitr)
options(width=60)
opts_chunk$set(cache=TRUE)
```





# Introduction

The package has been documented mostly with  Axion multi-well MEA
data, but it can work with any data.  Here we show how data can be
read in using a generic text reader, with a simple data format.  To
read in the data, we need two data files, and we need to provide
information about the layout of the plate.


```{r}
require(meaRtools)
show_top_file <- function(file) {
  cat(readLines(file, 10), sep='\n')
}
```

# Data format

The text reader requires two text files to be prepared: one with the
spike times and one with the channel positions.

## Spike times


A CSV file is used to store at least two columns: *Channel* containing
the channel name, and *Time* containing the time (in seconds) at which
a spike was detected on that channel.  The rows of the CSV do not need
to be ordered by channel, but the times for each channel should be
ordered, earliest spike first.

## Channel positions

A second CSV contains the (x,y) spatial location each channel.  If the
recording is from a multi-well array, then an extra *Well* column
determines the name of the well.  (If no Well column is provided, the
package assumes a plate with one well.)

## Plate information

The package by default has information about just two plates, the
Axion 12 and 48 well systems.  As shown in the examples below, in
addition we need to provide information regarding the plate.  If any
information is missing, the system will revert to defaults.

## Example 1: a single-well MEA recording

Here there is just a single MEA recording, from a hexagonal MEA (Wong
et al., 1993).  As there is only one well, the Well information is
absent from the position file.  The data files are provided within the
package, and the top of each file looks as follows:

```{r,comment=NA}
times = system.file("extdata/textreader/wong1993_p0.times", package="meaRtools")
pos = system.file("extdata/textreader/wong1993_p0.pos", package="meaRtools")
show_top_file(times)
show_top_file(pos)
```


```{r}
hex_platelayout = list(n_well = 1, #number of wells 
                        wells = c("w1"), #names of those wells.
                        n_well_r = 1, # number of wells / row
                        n_well_c = 1, # number of wells / col
                        layout = c(1, 1), # layout when plotting
                        n_elec_r = 8,
                        n_elec_c = 8,
                        xlim = c(-400, 400), # xlimits for plotting
                        ylim = c(-400, 400), # ylimits for plotting
                        spacing = 50,  # distance (um) separating electrodes
                        corr_breaks = 0 # vector of correlation distances
                        )

add_plateinfo("hex-1well", hex_platelayout)
s = read_spikelist_text(times, pos, array="hex-1well")
meaRtools:::.plot_mealayout(s$layout, use_names = TRUE, cex=0.3)
meaRtools:::.plot_meanfiringrate(s, main = "Mean Firing Rate by Plate (Hz)")
```




## Example 2: a synthetic multi-well recording

This second dataset is a composite of six recordings from P9 and P11
mouse retina (Demas et al, 2013), synthesised to make a 6-well plate.

```{r}
demas_platelayout = list(n_well = 6,
                        wells = paste0("w", 1:6),
                        n_well_r = 2,
                        n_well_c = 3,
                        layout = c(3, 2),
                        n_elec_r = 8,
                        n_elec_c = 8,
                        xlim = c(-100, 7200),
                        ylim = c(0, 6000),
                        spacing = 200,
                        corr_breaks = 0
                        )
add_plateinfo("demas-6well", demas_platelayout)
times = system.file("extdata/textreader/demas.times", package="meaRtools")
pos = system.file("extdata/textreader/demas.pos", package="meaRtools")
s = read_spikelist_text(times, pos, array="demas-6well")
meaRtools:::.plot_mealayout(s$layout, use_names = TRUE, cex=0.3)
```






# References

The test data come from the following two references:

Demas J, Eglen SJ, Wong ROL (2003) Developmental loss of synchronous spontaneous activity in the mouse retina is independent of visual experience. Journal of Neuroscience 23:2851–2860 Available at: <https://www.ncbi.nlm.nih.gov/pubmed/12684472>.

Wong RO, Meister M, Shatz CJ (1993) Transient period of correlated bursting activity during development of the mammalian retina. Neuron 11:923–938 <http://dx.doi.org/10.1016/0896-6273(93)90122-8>.

<!--This file can be rendered using rmarkdown::render("data_input.Rmd")   -->
